---
title: "Exploratory Data Analysis"
author: "John"
format: pdf
editor: visual
---

# Libraries and setup

```{r, message=FALSE, warning=FALSE}
library(here)
library(dplyr)
library(ggplot2)
library(ggforce)
source(here("R", "create_finaldata.R"))
```

Get the column variable names. Then call head() and tail() functions to take a quick look at the dataset.
```{r}
names(finaldata)
head(finaldata, 10)
tail(finaldata, 10)
```

# Missing Data Analysis
## Call summary to take a glance at the summary statistics of our dataset.
```{r}
finaldata |>
  summary()
```

## Count and proportion of NAs for each variable
```{r}
var <- c("country_name", "ISO", "region", "year", "gdp1000", "OECD", "OECD2023", 
               "popdens", "urban", "agedep", "male_edu", "temp", "rainfall1000", "matmor",
               "infmor", "neomor", "un5mor", "earthquake", "drought", "totdeath", "armconf1")

# Calculate the number and proportion of missing values for each variable
na_table <- data.frame(
  Variable = var,
  NA_Count = sapply(var, function(var) sum(is.na(finaldata[[var]]))),
  Proportion_NA = sapply(var, function(var) sum(is.na(finaldata[[var]])) / nrow(finaldata))
)

print(na_table)
```

Not sure which missing mechanism each variable is --MCAR, MAR, MNAR. I think it is best to ask some domain experts.
\newpage

# Visualizing matmor, infmor, neomor, un5mor,

```{r, message=FALSE, warning=FALSE}
grouped_data <- finaldata %>%
  group_by(region, year) %>%
  summarise(matmor = mean(matmor, na.rm = TRUE))  # Summarize matmor by region and year

ggplot(grouped_data, aes(x = year, y = matmor, group = region, color = region)) +
  geom_line() +
  geom_point() +
  labs(title = "Maternal Mortality Rate by Region Over Time",
       x = "Year",
       y = "Average Maternal Mortality Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")  # Move the legend to the bottom
```
\newpage

```{r, message=FALSE, warning=FALSE}
# Group by region and year, and summarize the neomor
grouped_data <- finaldata %>%
  group_by(region, year) %>%
  summarise(neomor = mean(neomor, na.rm = TRUE))  # Summarize neomor by region and year

# Create the plot
ggplot(grouped_data, aes(x = year, y = neomor, group = region, color = region)) +
  geom_line() +
  geom_point() +
  labs(title = "Neonatal Mortality Rate by Region Over Time",
       x = "Year",
       y = "Average Neonatal Mortality Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = 'bottom')  # Tilting x-axis labels for readability
```
\newpage

```{r, message=FALSE, warning=FALSE}
# Group by region and year, and summarize the infmor
grouped_data <- finaldata %>%
  group_by(region, year) %>%
  summarise(infmor = mean(infmor, na.rm = TRUE))  # Summarize infmor by region and year

# Create the plot
ggplot(grouped_data, aes(x = year, y = infmor, group = region, color = region)) +
  geom_line() +
  geom_point() +
  labs(title = "Infant Mortality Rate by Region Over Time",
       x = "Year",
       y = "Average Infant Mortality Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = 'bottom')  # Tilting x-axis labels for readability
```
\newpage

```{r, message=FALSE, warning=FALSE}
# Group by region and year, and summarize the un5mor
grouped_data <- finaldata %>%
  group_by(region, year) %>%
  summarise(un5mor = mean(un5mor, na.rm = TRUE))  # Summarize un5mor by region and year

# Create the plot
ggplot(grouped_data, aes(x = year, y = un5mor, group = region, color = region)) +
  geom_line() +
  geom_point() +
  labs(title = "Under 5 Mortality Rate by Region Over Time",
       x = "Year",
       y = "Average Under 5 Mortality Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = 'bottom')  # Tilting x-axis labels for readability
```

In general, all four mortality rates seems to be going down across all regions by year.
\newpage

Combine matmor, infmor, neomor, and un5mor to create a new dataset by country. 
```{r}
totalmort <- finaldata %>%
  group_by(country_name) %>%
  summarise(
    total_matmor = sum(matmor, na.rm = TRUE),
    total_infmor = sum(infmor, na.rm = TRUE),
    total_neomor = sum(neomor, na.rm = TRUE),
    total_un5mor = sum(un5mor, na.rm = TRUE)
  ) %>%
  mutate(totalmort = total_matmor + total_infmor + total_neomor + total_un5mor)
```




```{r, warning=FALSE, message=FALSE}
# Find the country with the maximum totalmort
max_country <- totalmort %>%
  filter(totalmort == max(totalmort, na.rm = TRUE)) %>%
  select(country_name, totalmort)

# Find the country with the minimum totalmort
min_country <- totalmort %>%
  filter(totalmort == min(totalmort, na.rm = TRUE)) %>%
  select(country_name, totalmort)

# View the results
print(max_country, max_country$totalmort)
print(min_country, min_country$totalmort)

```
The two countries with max and min total mortality rates are Sierra Leone with $34614.5$ deaths and Iceland with $221$ deaths.

```{r}
# Filter the data for the max and min countries and log-transform the mortality rates
combined_data <- finaldata %>%
  filter(country_name %in% c(max_country$country_name, min_country$country_name)) %>%
  select(country_name, matmor, infmor, neomor, un5mor) %>%
  mutate(
    log_matmor = log(matmor + 1),  # log transformation, adding 1 to avoid log(0)
    log_infmor = log(infmor + 1),
    log_neomor = log(neomor + 1),
    log_un5mor = log(un5mor + 1)
  ) %>%
  select(country_name, log_matmor, log_infmor, log_neomor, log_un5mor) %>%
  gather(key = "mortality_type", value = "log_mortality_rate", log_matmor, log_infmor, log_neomor, log_un5mor)

# Visualize the log-transformed data
ggplot(combined_data, aes(x = mortality_type, y = log_mortality_rate, fill = country_name)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Log-transformed Mortality Components for Countries with Max and Min Total Mortality",
       x = "Mortality Type",
       y = "Log-transformed Mortality Rate",
       fill = "Country") +
  theme_minimal()


```
Taking a look at the log transformed mortality rates for Sierra Leone and Iceland. 
Intepretation: If the y-axis value is $2$, the actual mortality rate is 
$\exp(2)−1 \approx 6.39$ per N deaths where N is the measurement metric from the dataset. 
\newpage

Now I will generate a random country and see if this country has similar rates to Sierra Leone and Iceland.
```{r, warning=FALSE, message=FALSE}
# Select a random country from the dataset
set.seed(123)  # Set seed for reproducibility
random_country <- finaldata %>%
  select(country_name) %>%
  distinct() %>%
  sample_n(1)  # Randomly select one country

# Specify Sierra Leone and Iceland
selected_countries <- c("Sierra Leone", "Iceland", random_country$country_name)

# Filter the data for the selected countries (random country, Sierra Leone, and Iceland)
combined_country_data <- finaldata %>%
  filter(country_name %in% selected_countries) %>%
  select(country_name, matmor, infmor, neomor, un5mor) %>%
  mutate(
    log_matmor = log(matmor + 1),  # log transformation, adding 1 to avoid log(0)
    log_infmor = log(infmor + 1),
    log_neomor = log(neomor + 1),
    log_un5mor = log(un5mor + 1)
  ) %>%
  select(country_name, log_matmor, log_infmor, log_neomor, log_un5mor) %>%
  gather(key = "mortality_type", value = "log_mortality_rate", log_matmor, log_infmor, log_neomor, log_un5mor)

# Visualize the log-transformed data for the selected countries
ggplot(combined_country_data, aes(x = mortality_type, y = log_mortality_rate, fill = country_name)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = paste("Log-transformed Mortality Components for", 
                     random_country$country_name, "Sierra Leone, and Iceland"),
       x = "Mortality Type",
       y = "Log-transformed Mortality Rate",
       fill = "Country") +
  theme_minimal()
```


Now look at Canada with Sierra Leone and Iceland.
```{r, warning=FALSE, message=FALSE}
# Specify Sierra Leone, Iceland, and Canada
selected_countries <- c("Sierra Leone", "Iceland", "Canada")

# Filter the data for the selected countries (Canada, Sierra Leone, and Iceland)
combined_country_data <- finaldata %>%
  filter(country_name %in% selected_countries) %>%
  select(country_name, matmor, infmor, neomor, un5mor) %>%
  mutate(
    log_matmor = log(matmor + 1),  # log transformation, adding 1 to avoid log(0)
    log_infmor = log(infmor + 1),
    log_neomor = log(neomor + 1),
    log_un5mor = log(un5mor + 1)
  ) %>%
  select(country_name, log_matmor, log_infmor, log_neomor, log_un5mor) %>%
  gather(key = "mortality_type", value = "log_mortality_rate", log_matmor, log_infmor, log_neomor, log_un5mor)

# Visualize the log-transformed data for the selected countries
ggplot(combined_country_data, aes(x = mortality_type, y = log_mortality_rate, fill = country_name)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Log-transformed Mortality Components for Canada, Sierra Leone, and Iceland",
       x = "Mortality Type",
       y = "Log-transformed Mortality Rate",
       fill = "Country") +
  theme_minimal()

```
# Summary
We looked at the column variable names, missing values for each variable and their proportions, visialized total mortality rate for the country with max/min mortality rates, and compared it to a random country and Canada. 